Generating Random Logic Programs Using Constraint Programming

### 2020-01-09 paper/paper.tex
Editing

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$. All constraint variables in the model are integer
variables, but, e.g., if the integer $i$ refers to a logical variable $X$, we
will use $i$ and $X$ interchangeably. All indices start at zero.

@@

\section{Considering Each Clause Separately}
% TODO: define heads & bodies

\subsection{Heads of Clauses}
% TODO: section for clauses. define heads & bodies, then separate subsections
% for head and body constraints

Our definition is slightly more involved because we want to eliminate some
symmetries.

\begin{definition}
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$.
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}
    \cup \{ \Box \}$
  \end{itemize}
  In both cases, $\Box$ is the disabled value.
\end{definition}

\begin{definition}
  The \variable{predicate}'s $\variable{arity} \in [0, \maxArity{}]$ can then be
  defined using the \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      0 & \text{if } \variable{predicate} = \Box\\
      \text{the arity of } \variable{predicate} & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    \variable{arguments}[i] = \Box \iff i \ge \variable{arity}.
  \]
\end{constraint}

@@

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{} \cup \{ \Box
    \}}$, where $\Box$ denotes a disabled argument (the reason why this value is
  needed will become clear in \cref{sec:variable_symmetry}). The node's
  $\variable{arity} \in [0, \maxArity{}]$ is defined by a \variable{table}
  constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{name} & \text{if } \variable{name} \in
      \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
  \variable{arguments}[i] = \Box \quad \iff \quad i \ge \variable{arity}.
  \]
\end{constraint}
% TODO: this could merged into a single constraint for termsPerClause

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{align*}
    \variable{count}(i, \variable{treeStructure}_{-i}) = 0 &\iff \variable{treeValues}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    \variable{count}(i, \variable{treeStructure}_{-i}) = 1 &\iff \variable{treeValues}[i].\variable{name} = \neg,\\
    \variable{count}(i, \variable{treeStructure}_{-i}) > 1 &\iff \variable{treeValues}[i].\variable{name} \in \{ \land, \lor \}.
  \end{align*}
  $\variable{treeStructure}_{-i}$ denotes array $\variable{treeStructure}$ with
  position $i$ skipped.
\end{constraint}
Each constraint corresponds to node $i$ having no children, one child, and
multiple children, respectively.

@@

\begin{definition}
  % TODO: what odes this mean? example?
  Let $N = \maxArity{} \times (\maxNumNodes{} + 1)$. Let
  $\variable{termsPerClause}[N] \in \constants{} \cup \variables{} \cup \{ \Box
  \}$ be a matrix, each row of which is a flattened list of all positions for
  terms in a particular clause.

  Then we can use the $\variable{setsIntsChanneling}$ constraint
  to define $\variable{occurrences}[|\constants{}| + |\variables{}| + 1]$ as a
  matrix of subsets of $\{ 0, \dots, N-1 \}$ such that for all $i = 0, \dots, N
  - 1$, and $t \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    i \in \variable{occurrences}[t] \quad \iff \quad
    \variable{termsPerClause}[i] = t
  \]
\end{definition}

\begin{definition}
  We define $\arrayd{M}{|\variables{}|}{\{ 0, \dots, N \}}$ such that for $v \in
  \variables{}$,
  \[
    \variable{M}[v] = \begin{cases}
      \min \variable{occurrences}[v] & \text{if }
      \variable{occurrences}[v] \ne \emptyset\\
      N & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  $\variable{M}$ is sorted.
\end{constraint}

\begin{example} % TODO: update the example to cover an entire clause
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\constants{} = \{ a
  \}$, $\variables{} = \{X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be
  represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

\section{Counting Programs}
% TODO: define combined arity.
% TODO: define 'gap', 'argument', 'position' or something like that.

In order to demonstrate the correctness of the model and explain it in more
detail, in this section we are going to derive combinatorial expressions for
counting the number of programs with up to $\maxNumClauses{}$ clauses and up to
$\maxNumNodes{}$ nodes per clause, and arbitrary $\predicates{}$,
$\arities{}$, $\variables{}$, and $\constants{}$. To simplify the task, we only
consider clauses without probabilities and disable (negative) cycle elimination.
It was experimentally confirmed that the model agrees with the combinatorial
formula from this section in 985 different scenarios.

We will first consider clauses with `gaps', i.e., without taking variables and
constants into account. Let $T(n, a)$ denote the number of possible clause
bodies with $n$ nodes and combined arity $a$. Then $T(1, a)$ is the number of
predicates in $\predicates{}$ with arity $a$, and the following recursive
definition can be applied for $n > 1$:
\[
  T(n, a) = T(n-1, a) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
      c_i \ge 1 \text{ for all } i}} \sum_{\substack{d_1 + \dots + d_k = a,\\
    d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\]
The first term here represents negation, i.e., negating an expression consumes
one node but otherwise leaves the task unchanged. If the first operation is not
negation, then it must be either conjunction or disjunction (hence the
coefficient `2'). In the first sum, $k$ represents the number of children of the
root node, and each $c_i$ is the number of nodes dedicated to child $i$. Thus,
the first sum iterates over all possible ways to partition the remaining $n-1$
nodes. Similarly, the second sum considers every possible way to partition the
total arity $a$ across the $k$ children nodes.

@@

The number of ways to fill $n$ positions with terms can be expressed as
\[
  P(n) = |\constants{}|^n + \sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}} \prod_{i=0}^k (|\constants{}| +
  i)^{s_{i+1} - s_i - 1}.
\]
The first term is simply the number of ways to fill all $n$ positions with
constants. The parameter $k$ is the number of variables used in the clause, and
$s_1, \dots, s_k$ mark the first occurrence of each variable. For each gap
between any two introductions (or before the first introduction, or after the last
introduction), we have $s_{i+1}-s_i-1$ spaces to be filled with any of the
$\constants{}$ constants or any of the $i$ already-introduced variables.

>>>>

% TODO: define 'gap'

We will often use $\Box$ as a special domain value indicating a `disabled'
(i.e., fixed and ignored) part of the model. We write $\arrayd{a}{b}{c}$ to mean
that $\variable{a}$ is an array of variables of length $b$ such that each
element of $\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$
to denote an array $\variable{a}$ of length $b$ such that each element of
$\variable{a}$ has type $\variable{c}$. All constraint variables in the model
are integer variables, but, e.g., if the integer $i$ refers to a logical
variable $X$, we will use $i$ and $X$ interchangeably. All indices start at
zero.

@@

\section{Considering Each Clause Separately}

\subsection{Heads of Clauses}

\begin{definition}
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$.
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}
    \cup \{ \Box \}$
  \end{itemize}
\end{definition}
The reason why $\Box$ must be a separate value will become clear in
\cref{sec:variable_symmetry}.

\begin{definition} \label{def:arity}
  The \variable{predicate}'s $\variable{arity} \in [0, \maxArity{}]$ can then be
  defined using the \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{predicate} & \text{if }
      \variable{predicate} \in \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint} \label{constr:arity}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    \variable{arguments}[i] = \Box \iff i \ge \variable{arity}.
  \]
\end{constraint}

@@

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{} \cup \{ \Box
    \}}$. The node's $\variable{arity}$ can then be
  defined analogously to \cref{def:arity}.
\end{definition}

We can use \cref{constr:arity} again to disable extra arguments.

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{align*}
    \variable{count}(i, \variable{treeStructure}_{-i}) = 0 &\iff \variable{treeValues}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    \variable{count}(i, \variable{treeStructure}_{-i}) = 1 &\iff \variable{treeValues}[i].\variable{name} = \neg,\\
    \variable{count}(i, \variable{treeStructure}_{-i}) > 1 &\iff \variable{treeValues}[i].\variable{name} \in \{ \land, \lor \},
  \end{align*}
  where $\variable{treeStructure}_{-i}$ denotes array $\variable{treeStructure}$
  with position $i$ skipped.
\end{constraint}
Each constraint corresponds to node $i$ having no children, one child, and
multiple children, respectively.

@@

\begin{definition}
  Let $N = \maxArity{} \times (\maxNumNodes{} + 1)$. Let
  $\variable{terms}[N] \in \constants{} \cup \variables{} \cup \{ \Box
  \}$ be a flattened array of all gaps in a particular clause.

  Then we can use the $\variable{setsIntsChanneling}$ constraint
  to define $\variable{occurrences}[|\constants{}| + |\variables{}| + 1]$ as an
  array of subsets of $\{ 0, \dots, N-1 \}$ such that for all $i = 0, \dots, N
  - 1$, and $t \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    i \in \variable{occurrences}[t] \quad \iff \quad
    \variable{terms}[i] = t
  \]
\end{definition}

\begin{definition}
  We define $\arrayd{introductions}{|\variables{}|}{\{ 0, \dots, N \}}$ such
  that for $v \in \variables{}$,
  \[
    \variable{introductions}[v] = \begin{cases}
      \min \variable{occurrences}[v] & \text{if }
      \variable{occurrences}[v] \ne \emptyset\\
      N & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  $\variable{introductions}$ are sorted.
\end{constraint}

\begin{example}
  Let $\constants{} = \emptyset$, $\variables{} = \{ X, Y, Z \}$, $\maxArity{} =
  2$, $\maxNumNodes{} = 3$, and consider the clause
  \[
    \variable{sibling}(X, Y) \gets \variable{parent}(X, Z) \land
    \variable{parent}(Y, Z).
  \]
  Then $\variable{terms} = [X, Y, \Box, \Box, X, Z, Y, Z]$ (the boxes represent
  the conjunction node), $\variable{occurrences} = [\{ 0, 4 \}, \{ 1, 6 \},
  \{ 5, 7 \}, \{ 2, 3 \}]$, and $\variable{introductions} = [0, 1, 5]$.
\end{example}

@@

\section{Counting Programs}

In order to demonstrate the correctness of the model and explain it in more
detail, in this section we are going to derive combinatorial expressions for
counting the number of programs with up to $\maxNumClauses{}$ clauses and up to
$\maxNumNodes{}$ nodes per clause, and arbitrary $\predicates{}$,
$\arities{}$, $\variables{}$, and $\constants{}$. To simplify the task, we only
consider clauses without probabilities and disable (negative) cycle elimination.
It was experimentally confirmed that the model agrees with the combinatorial
formula from this section in 985 different scenarios. The \emph{total arity} of
a body of a clause is the sum total of arities of all predicates in the body.

We will first consider clauses with gaps, i.e., without taking variables and
constants into account. Let $T(n, a)$ denote the number of possible clause
bodies with $n$ nodes and total arity $a$. Then $T(1, a)$ is the number of
predicates in $\predicates{}$ with arity $a$, and the following recursive
definition can be applied for $n > 1$:
\[
  T(n, a) = T(n-1, a) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
      c_i \ge 1 \text{ for all } i}} \sum_{\substack{d_1 + \dots + d_k = a,\\
    d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\]
The first term here represents negation, i.e., negating an expression consumes
one node but otherwise leaves the task unchanged. If the first operation is not
negation, then it must be either conjunction or disjunction (hence the
coefficient `2'). In the first sum, $k$ represents the number of children of the
root node, and each $c_i$ is the number of nodes dedicated to child $i$. Thus,
the first sum iterates over all possible ways to partition the remaining $n-1$
nodes. Similarly, the second sum considers every possible way to partition the
total arity $a$ across the $k$ children nodes.

@@

The number of ways to fill $n$ gaps with terms can be expressed as
\[
  P(n) = |\constants{}|^n + \sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}} \prod_{i=0}^k (|\constants{}| +
  i)^{s_{i+1} - s_i - 1}.
\]
The first term is simply the number of ways to fill all $n$ gaps with
constants. The parameter $k$ is the number of variables used in the clause, and
$s_1, \dots, s_k$ mark the first occurrence of each variable. For each gap
between any two introductions (or before the first introduction, or after the last
introduction), we have $s_{i+1}-s_i-1$ spaces to be filled with any of the
$\constants{}$ constants or any of the $i$ already-introduced variables.
